---
title: "Hammell_Analysis_OTHER"
author: "Abbey Hammell"
output: html_document
---

The following code outlines the Pre-acqusition and Acquisition phase analyses completed for Hammell et al., "Temporal Course of Over-Generalized Conditioned Threat Expectancies in Posttraumatic Stress Disorder". Note that this document is meant for reproducibility/replicability purposes for Pre-acqusition and Acquisition phases. To understand the general procedure for the SSANOVA analyses and contrast analyses, it is recommended that you review the "Hammell_Analysis_GEN.rmd" file, which contains a better walkthrough/outline of the procedure. Please note that the groups in our PreACQ and ACQ data were specified as 2 = Trauma Control; 3 = SubPTSD; and 4 = PTSD. 

#Set up R workspace
```{r}
rm(list = ls()) #clear workspace

pacman::p_load(bigsplines) #install & attach bigsplines package (note: the "pacman" package needs to be installed)
```



#PRE-ACQUISITION ANALYSES

###Load Data
```{r load-data}
datapath <- "~/Desktop/Quick Access Files/" #set directory of where data is located
dataPREACQ <- "PreACQData_201901.csv" #give name of datafile
ptsd.PreACQ <- read.csv(file = paste0(datapath, dataPREACQ),
                 header = TRUE) #load data
```

### Convert to Data Frame
```{r convert-data-frame}
slevels <- c("vCS-", "oCS-", "GS1", "GS2", "GS3", "CS+")
nvar <- 6L 
ptsd.PreACQ.df <- data.frame(SubID = rep(ptsd.PreACQ$ID, each = nvar),
                      Group = rep(ptsd.PreACQ$Group, each = nvar),
                      Stimulus = rep(slevels, each = 1),
                      Risk = c(t(ptsd.PreACQ[,4:9])))
```

### Convert Factors
```{r convert-factors}
ptsd.PreACQ.df$SubID <- factor(ptsd.PreACQ.df$SubID) #ensure subject ID is a factor variable
ptsd.PreACQ.df$Stimulus <- factor(ptsd.PreACQ.df$Stimulus,
                           levels = slevels, ordered = TRUE) #ensure Stimulus is an ordered factor
ptsd.PreACQ.df$Group <- factor(ptsd.PreACQ.df$Group) #ensure group is a factor variable
```

IMPORTANT NOTE: N/As weren't taken out with code for PREACQ because there were no missing observations for Risk. You will want to take out missing observations at this point if you do have any. 


## Fitting the SSANOVA Model 

### Get Knots
Our overall nonparametric regression model was fit using the "bigssp" function in the “bigsplines” R package. Before we fit the model, we must specify the number and location of our knots. The knots are the locations across the values of the independent/predictor variables at which the model coefficients are allowed to change. Below is the code that we used to obtain our knots for the Acquisition phase. All unique predictor value combinations were used as knots. The present model uses 180 knots (there are 3 groups x 6 stimuli, so there is a total of 18 unique values of y to estimate).

```{r}
#we do not include trials in this code like we do for GEN or ACQ because we are not interested in trials
knotID <- NULL
  for(jj in levels(ptsd.PreACQ.df$Stimulus)){ #for jj in levels of stimulus
    idS <- (ptsd.PreACQ.df$Stimulus == jj)
    for(kk in levels(ptsd.PreACQ.df$Group)){ #for kk in levels of group
      idG <- (ptsd.PreACQ.df$Group == kk)
      ixnew <- which(idS & idG)
      knotID <- c(knotID, ixnew[1])
    }
  }
```

### Fit Model  
```{r fit-model}
types <- list(Group = "nom", Stimulus = "ord") #specify what types of spline each predictor variable will use
rmod <- bigssp(Risk ~ Group * Stimulus,
               data = ptsd.PreACQ.df,
               nknots = knotID,
               type = types,
               random = ~ Group | SubID,
               skip.iter = F)
```

### Variance Parameters
```{r variance-parameters}
rmod$sigma^2 * c(error = 1, rmod$tau)
```

### Intraclass Correlation Coefficient (ICC)
```{r icc}
rmod$tau / (1 + rmod$tau)
```

### R-squared Values
```{r r-squared}
blupID <- match(ptsd.PreACQ.df$SubID, names(rmod$blup))
blups <- rmod$blup[blupID]
yhat <- rmod$fitted.values + blups
cor(ptsd.PreACQ.df$Risk, rmod$fitted.values)^2  # fixed effects only
cor(ptsd.PreACQ.df$Risk, yhat)^2  # fixed and random effects
```

### Histogram of the residuals
```{r}
hist(ptsd.PreACQ.df$Risk - yhat)
```






# ACQUISITION ANALYSES 

## STEP 1 - Preparing the Data for Analyses

### Load Data
```{r}
datapath <- "~/Desktop/Quick Access Files/" #set directory of where data is located
dataACQ <- "MARSRR_ACQData.csv" #give name of datafile
ptsd.ACQ <- read.csv(file=paste0(datapath, dataACQ), header=T)
```

### Rename Variables
```{r}
vnames <- c("SubID","Group", 
            paste(paste0("P",1:8), rep(c("vCSm","oCSm","CSp"),each=8),sep="."))
names(ptsd.ACQ) <- vnames
```

### Convert to Data Frame
```{r}
nvar <- 24L #3 trials x 8 stimuli
ptsd.ACQ.df <- data.frame(SubID = rep(ptsd.ACQ$SubID, each=24),
                      Group = rep(ptsd.ACQ$Group, each=24),
                      Trial = rep(1:8, 3),
                      Stimulus = rep(c("vCS-","oCS-","CS+"),each=8),
                      Risk = c(t(ptsd.ACQ[,-c(1:2)])))
```

###Convert Factors
```{r}
ptsd.ACQ.df$SubID <- factor(ptsd.ACQ.df$SubID) #ensure subject ID is a factor variable
ptsd.ACQ.df$Group <- factor(ptsd.ACQ.df$Group, ordered = FALSE) #ordered = F specified because our groups were specified by numbers originally, but this was not the same for generalization 
ptsd.ACQ.df$Stimulus <- factor(ptsd.ACQ.df$Stimulus, ordered=TRUE, levels=c("vCS-","oCS-","CS+")) #convert stimulus to ordered factors
```

###Remove NA Data
```{r}
ix <- which(is.na(ptsd.ACQ.df$Risk)) #find which observations are missing
ptsd.ACQ.df <- ptsd.ACQ.df[-ix,] #remove missing observations from the dataset 
```

## STEP 2 - Fitting the SSANOVA Model

### Get knots
```{r}
#note that this for loop does the same thing as in the GEN script; it's just an alternative way of coding the for loop
knotID <- NULL 
for(ii in 1:8){ #for ii in number of trial (8 trials in Acquisition)
  for(jj in levels(ptsd.ACQ.df$Stimulus)){
    for(kk in levels(ptsd.ACQ.df$Group)){
      ixnew <- which(ptsd.ACQ.df$Trial==ii & ptsd.ACQ.df$Stimulus==jj & ptsd.ACQ.df$Group==kk)
      knotID <- c(knotID, ixnew[1])
    }
  }
}
```

### Fit model
```{r}
types <- list(Trial="cub",Stimulus="ord",Group="nom") #specify what types of spline each IV will use
rmod.ACQ <- bigssp(Risk ~ Trial * Group * Stimulus, 
                   data=ptsd.ACQ.df, 
                   nknots=knotID, 
                   type= types, 
                   random=  ~ Group | SubID, 
                   skip.iter=T)
```

## Model Diagnostics 

### Variance Parameters
```{r}
#obtaining variance parameters of the random effects
rmod.ACQ$sigma^2 * c(error = 1, rmod.ACQ$tau)
```

### Intraclass Correlation Coefficient (ICC)
```{r}
#obtaining intraclass correlation coefficients
rmod.ACQ$tau / (1 + rmod.ACQ$tau)
```

### R-squared
```{r}
yhat <- rmod.ACQ$fitted.values + rmod.ACQ$blup[match(ptsd.ACQ.df$SubID,names(rmod.ACQ$blup))]
cor(ptsd.ACQ.df$Risk, rmod.ACQ$fitted.values)^2 #fixed effects only
cor(ptsd.ACQ.df$Risk, yhat)^2 #fixed and random effects
```

### Histogram of the residuals
```{r}
hist(ptsd.ACQ.df$Risk - yhat)
```

### Cosine diagnostics
```{r}
summary(rmod.ACQ, diagnostics = TRUE)$pi
```


## STEP 3 - Stimulus Contrasts

R code to get differences between estimated CS+ and oCS- response, relative to the vCS- response, for each separate group. Note that the code for the stimulus contrasts and group-stimulus contrasts for the Acquisition phase are almost exactly the same as the code used for analyzing the Generalization phase (see other R markdown file associated with this paper). The only modifications made are those to accomodate the change in trial number. We also calculated 95% CIs to put into the completed dataframe.

First, we find the matrix square-root of the covariance matrix of the coefficients, which will be used for calculation of the standard errors of our contrasts. Next, we want to specify our reference stimulus (vCS) and specify a vector of stimuli ("oth") that we are contrasting with our reference stimulus.
```{r}
coefsqrt <- rmod.ACQ$modelspec$coef.csqrt
ref <- "vCS-"  #this can be changed to "oCS-" if you wanted to, say, contrast the CS+ with the oCS-
oth <- levels(ptsd.ACQ.df$Stimulus)[-1] # in our case, assigning "oth" as a vector of our other two stimuli, "CS+" and "oCS-"
```

Next, we can run the for-loop below that calculates stimulus contrasts for each group at each stimulus/trial combination. 
```{r stimulus contrasts}
difStm.ACQ <- NULL
for(g in levels(ptsd.ACQ.df$Group)){
  newdata.ref <- expand.grid(Trial = 1:8, #we have 8 risk trials per stimulus type
                             Group = g,
                             Stimulus = ref)
  pred.ref <- predict(rmod.ACQ, newdata.ref, design = TRUE)
  for(s in oth){
    newdata.oth <- expand.grid(Trial = 1:8, #we have 8 risk trials per stimulus type
                               Group = g,
                               Stimulus = s) 
    pred.oth <- predict(rmod.ACQ, newdata.oth, design = TRUE)
    dif.fit <- pred.oth$fit - pred.ref$fit
    dif.X <- pred.oth$X - pred.ref$X
    dif.var <- rowSums( (dif.X %*% coefsqrt)^2 )
    dif.se <- sqrt(dif.var)
    Tstat <- dif.fit / dif.se
    CI95 <- paste0("[",round(dif.fit - (1.96*dif.se),2),", ",round(dif.fit + (1.96*dif.se),2),"]")
    pval <- 2*(1 - pnorm(abs(Tstat)))
    newdiff <- data.frame(fit = dif.fit, #estimates
                          se.fit = dif.se, #standard error of the estimates
                          t.value = Tstat, #t-stat
                          p.value = pval, #p-value
                          CI.95 = CI95, #upper and lower 95% confidence interval
                          Trial = 1:8, 
                          Group = g,
                          Stimulus = s,
                          Reference = ref) #reference stimulus
    difStm.ACQ <- rbind(difStm.ACQ, newdiff)
  }
}
head(difStm.ACQ)
```

## STEP 4 - Group-Stimulus Contrasts

```{r}
coefsqrt <- rmod.ACQ$modelspec$coef.csqrt
glevels <- levels(ptsd.ACQ.df$Group)
ref <- "oCS-"   # reference
oth <- "CS+"  # other stimulus
#Note, you can change the "oth" stimulus and "ref" stimulus to any combination you want and re-run the for-loop below to get results for a contrast between any specific stimuli of your choosing; recall that the "ref" stimulus will always have its estimate subtracted from the "oth" stimulus' estimate. In the code above, we are specifying a CS+ vs. oCS- contrast. 
```

```{r stimulus-group-contrasts}
difStmGrp.ACQ <- NULL
for(s in oth){ #for each stimulus specified in "oth"
  for(g1 in 1:2){ #for each Group in levels(ptsd.df$Group) index 1 & 2 (i.e., SubPTSD and TC groups)
    # Find stimulus contrasts for each reference group (SubPTSD or TC)
    newdata.ref1 <- expand.grid(Trial = 1:8,  #we have 8 risk trials per stimulus type
                                Group = glevels[g1],
                                Stimulus = ref)
    #predict values for reference (vCS-) for the 1st & 2nd Group in glevels
    pred.ref1 <- predict(rmod.ACQ, newdata.ref1, design = TRUE) 
    newdata.oth1 <- expand.grid(Trial = 1:8, #we have 8 risk trials per stimulus type
                                Group = glevels[g1],
                                Stimulus = s)
    #predict values for other stimuli for the 1st & 2nd Group in glevels
    pred.oth1 <- predict(rmod.ACQ, newdata.oth1, design = TRUE)
    #find the stimulus contrast for the 1st group
    dif.fit1 <- pred.oth1$fit - pred.ref1$fit
    dif.X1 <- pred.oth1$X - pred.ref1$X

    for(g2 in (g1+1):3){ #for each Group in levels(ptsd.df$Group) index 2 & 3 (i.e., PTSD and SubPTSD groups)

      # Find stimulus contrast for each comparison group (PTSD or SubPTSD)
      newdata.ref2 <- expand.grid(Trial = 1:8,  #we have 8 risk trials per stimulus type
                                  Group = glevels[g2],
                                  Stimulus = ref)
      pred.ref2 <- predict(rmod.ACQ, newdata.ref2, design = TRUE)
      newdata.oth2 <- expand.grid(Trial = 1:8,  #we have 8 risk trials per stimulus type
                                  Group = glevels[g2],
                                  Stimulus = s)
      pred.oth2 <- predict(rmod.ACQ, newdata.oth2, design = TRUE)
      dif.fit2 <- pred.oth2$fit - pred.ref2$fit
      dif.X2 <- pred.oth2$X - pred.ref2$X

      # Difference = Group 2 minus Group 1
      dif.fit <- dif.fit2 - dif.fit1
      dif.X <- dif.X2 - dif.X1
      dif.var <- rowSums( (dif.X %*% coefsqrt)^2 )
      dif.se <- sqrt(dif.var)
      CI95 <- paste0("[",round(dif.fit - (1.96*dif.se),2),", ",round(dif.fit + (1.96*dif.se),2),"]")
      Tval <- dif.fit / dif.se
      pval <- 2*(1 - pnorm(abs(Tval)))
      newdiff <- data.frame(fit = dif.fit, #estimates
                            se.fit = dif.se, #standard error of the estimates
                            t.value = Tval, #t-stat
                            p.value = pval, #p-value
                            Trial = 1:8, #there are 8 trials of risk appraisals per stimulus in the Acquisition phase
                            CI.95 = CI95, #upper and lower 95% confidence interval
                            Group1 = glevels[g1], #specifies first group
                            Group2 = glevels[g2], #specifies second group
                            Stimulus = s,
                            Reference = ref) #reference stimulus
      difStmGrp.ACQ <- rbind(difStmGrp.ACQ, newdiff)

    }
  } 
} 
head(difStmGrp.ACQ)
```

